Reagents: expressing and composing fine-grained concurrency

### 2011-11-06 paper/paper.tex
writing on implementation and dynamic reagents

@@

\begin{figure}
\begin{lstlisting}
class TreiberStack[A] {
  private val head = new AtomicReference[List[A]](Nil)
  def push(a: A) {
    val backoff = new Backoff
    while (true) {
      val cur = head.get()
      if (head.cas(cur, a :: cur)) return
      backoff.once()
    }
  }
  def tryPop(): Option[A] = {
    val backoff = new Backoff
    while (true) {
      val cur = head.get() 
      cur match {
        case Nil     => return None
        case a::tail => 
          if (head.cas(cur, tail)) return Some(a)
      }
      backoff.once()
    }
  }
}
\end{lstlisting}
\nocaptionrule
\caption{Treiber's stack (in Scala)}
\label{fig:classic-treiber}
\end{figure}

@@

\elide{
In \secref{background}, we saw a range of strategies for communication and
synchronizion at a fine grain:
\begin{itemize}
\item using CAS to make direct, optimistic, and atomic updates;
\item using exponential backoff to ease contention; 
\item using side-channel communication to increase parallelism; 
\item using disjoint spinwaiting to eliminate useless cache traffic; 
\item using dual data structures to incorporate blocking.
\end{itemize}
Reagents incorporate these disparate techniques into a uniform set of
combinators for writing concurrent algorithms.  Some of the techniques are
built into the combinators themselves, while others are expressible as
reusable abstractions.
}

@@

// combinators for composition
+  : Reagent[A,B] * Reagent[A,B] => Reagent[A,B]
>> : Reagent[A,B] * Reagent[B,C] => Reagent[A,C]
|||  : Reagent[A,B] * Reagent[A,C] => Reageng[A,(B,C)]

// combinators for computation
ret: A => Reagent[Unit, A]
computed: PartialFunction[A, Reagent[Unit, B]]
          => Reagent[A,B]
postCommit: (A => Unit) => Reagent[A,A]
\end{lstlisting}
\label{fig:combinators}
\end{figure}

@@

Memory is shared between reagents using the type \lstinline{Ref[A]} of
atomically-updatable references.\footnote{ These references are only
  accessible through reagent combinators, which allows them to provide
  functionality beyond the built-in \lstinline{AtomicReference[A]} type, such
  as the blocking logic we will see shortly.  }
The \lstinline{upd} combinator~(\figref{combinators}) represents simple atomic
updates to references.  It takes an \emph{update function}, which tells how to
transform a snapshot of the reference cell and some input into an updated
value for the cell and some output.  Using \lstinline{upd}, we can rewrite
\lstinline{TreiberStack} in a more readable and concise way:
\begin{lstlisting}
class TreiberStack[A] {
  private val head = new Ref[List[A]](Nil)
  val push: Reagent[A, Unit] = upd(head) {
    case (xs, x) => (x:xs, ())
  }
  val tryPop: Reagent[Unit, Option[A]] = upd(head) {
    case (x::xs, ()) => (xs, Some(x))
    case (Nil,   ()) => (Nil, None)
  }
}
\end{lstlisting}
In Scala, anonymous partial functions are written as a series of cases
enclosed in braces.  For \lstinline{push} and \lstinline{tryPop}, the case
analyses are exhaustive, so the update functions are in fact total.

@@

The \lstinline{!} operator invokes a reagent as a reactant: the invocation
lasts for a single reaction, and any messages sent by the reagent are
consumed.  But sometimes it is useful for reagents to persist beyond a single
reaction, \emph{i.e.}, to act as catalysts.  For example, the following
function creates a catalyst that merges input from two endpoints and sends the
resulting pairs to another endpoint:
\begin{lstlisting}
def zip(in1: Endpoint[Unit,  A], 
        in2: Endpoint[Unit,  B], 
        out: Endpoint[(A,B), Unit]) = 
  dissolve((swap(in1) ||| swap(in2)) >> swap(out))
\end{lstlisting}
The \lstinline{dissolve} function takes a \lstinline{Reagent[Unit, Unit]} and
introduces it as a catalyst.\footnote{For simplicity, we have not given a way
  to \emph{cancel} catalysts after they have been introduced, but cancellation
  is easy to add.}  Operationally, in this example, that just means sending
messages along \lstinline{in1} and \listinline{in2} that are marked as
catalyzing messages, and hence are not consumed during reaction.  The upshot
is that senders along \lstinline{in1} will see the catalyzing messages, look
for messages along \lstinline{in2} to pair with, and ultimately send messages
along \lstinline{out} (and similarly in the other order).

@@

\begin{figure}
\begin{lstlisting}
class MSQueue[A] {
  private case class Node(data: A, next: Ref[Node])
  private val head = new Ref(Node(null))
  private val tail = new Ref(head.read ! ())
  val tryDeq: Reagent[Unit, Option[A]] = upd(head) {
    case (Node(_, Ref(n@Node(x, _))), ()) => (n, Some(x))
    case (emp,                        ()) => (emp, None)
  }
  private def findAndEnq(n: Node): Reagent[Unit,Unit] = 
    read(tail) ! () match {
      case ov@Node(_, r@Ref(null)) => 
        cas(r, null, n) >>
        postCommit { _ => cas(tail, ov, n) !? () }
      case ov@Node(_, Ref(nv)) => 
        cas(tail, ov, nv) !? (); 
        findAndEnq(n)
    }
  val enq: Reagent[A, Unit] = computed { 
    (x: A) => findAndEnq(Node(x, new Ref(null)))
  }
}
\end{lstlisting}
\nocaptionrule
\caption{The Michael-Scott queue, with reagents}
\label{fig:msqueue}
\end{figure}

An example of this phenomenon is the classic Michael-Scott lock-free
queue~\cite{?}.

@@

\elide{
The \lstinline{read} combinator has a straightforward signature: if
\lstinline{r} has type \lstinline{Ref[A]}, then \lstinline{read(r)} is a
\emph{value} of type \lstinline{Reagent[Unit, A]} which, when invoked, takes a
unit\footnote{\lstinline{Unit} in Scala is akin to \lstinline{void}: it is a
  type with a single member, written \lstinline{()}.} argument and yields an
\lstinline{A} result.

The \lstinline{cas} combinator takes two additional \lstinline{A} arguments,
giving the expected and updated values, respectively.  Unlike its counterpart
for \lstinline{AtomicReference}, a \lstinline{cas} reagent does \emph{not}
yield a boolean result.  Instead, the reagent library builds in handling of
failed CASes, including an optimistic retry loop and exponential backoff.  The
failure mechanics depend on which other combinators are involved, so we delay
a complete discussion until \secref{failure}.
}

% tail recursive

@@

\paragraph{A caveat for composition}

@@

@@

@@

In the implementation, \lstinline{Reagent[A,B]} is an abstract class all of
whose subclasses are hidden.  The subclasses roughly correspond to the
combinator functions (which are responsible for instantiating them), and
instances of the subclasses store the arguments given to their corresponding
combinator.  Each subclass provides an implementation of the
\lstinline{tryReact} method, which is an abstract method of
\lstinline{Reagent[A,B]} with the following signature:\footnote{ Functional
  programming afficianados should regard \lstinline{tryReact} as a kind of
  Kleisli arrow for reagents.}
\begin{lstlisting}
  def tryReact(a: A, rx: Reaction, offer: Offer[B]): Any
\end{lstlisting}
The \lstinline{Any} type in Scala lies at the top of the subtyping hierarchy,
akin to \lstinline{Object} in Java.  Here we are using \lstinline{Any} to
represent a union of the type \lstinline{B} with the type
\lstinline{Failure}, where the latter has just two singleton instances,
\lstinline{Block} and \lstinline{Retry}, corresponding to permanent and
transient failures.

@@

\subsection{The entry point: reacting}
\label{sec:impl-reaction}

\begin{figure}
\begin{lstlisting}
def !(a: A): B = {
  val backoff = new Backoff
  def withoutOffer(): B =
    tryReact(a, empty, null) match {
      case Block => withOffer()
      case Retry => 
        backoff.once()
        if (maySync) withOffer() else withoutOffer()
      case ans   => ans.asInstanceOf[B]
    }
  def withOffer(): B = {
    val offer = new Offer[B]
    tryReact(a, empty, offer) match {
      case (f: Failure) => 
        if (f == Block) park() else backoff.once()
        if (offer.rescind) withOffer() else offer.answer
      case ans => ans.asInstanceOf[B]
    }
  }
  withoutOffer()
}
\end{lstlisting}
\nocaptionrule
\caption{The \lstinline{!} method, defined in \lstinline{Reagent[A,B]}}
\label{fig:reaction}
\end{figure}
%TODO: note tail recursion

@@

\begin{figure}
\begin{lstlisting}[frame=single]
class Commit[A] extends Reagent[A,A] {
  def tryReact(a: A, rx: Reaction, offer: Offer[A]) = 
    if (offer != null && !offer.rescind) offer.answer
    else if (rx.commit) a 
    else Retry
}
\end{lstlisting}
\vskip -12pt
\begin{lstlisting}[frame=single]
class CAS[A](ref: Ref[A], oldV: A, newV: A) 
  extends Reagent[Unit,Unit] {
  def tryReact(u: Unit, rx: Reaction, offer: Offer[B]) = 
    if (!rx.hasCAS && !k.hasCAS)
      if (ref.cas(oldV, newV))
        k.tryReact((), rx, offer)
      else Retry
    else
      k.tryReact((), rx.withCAS(ref, oldV, newV), offer)
}
\end{lstlisting}
\vskip -12pt
\begin{lstlisting}[frame=single]
class Choice[A,B](r1: Reagent[A,B], r2: Reagent[A,B]) 
  extends Reagent[A,B] {
  def tryReact(a: A, rx: Reaction, offer: Offer[B]) = 
    r1.tryReact(a, rx, offer) match {
    case Retry => 
        r2.tryReact(a, rx, offer) match {
          case Retry => Retry
          case Block => Retry // must still retry r1
          case ans   => ans
        }
      case Block => r2.tryReact(a, rx, offer)
      case ans => ans
    }
}
\end{lstlisting}
\vskip -12pt
\begin{lstlisting}[frame=single]
class Computed[A,B](c: A => Reagent[Unit,B]) 
  extends Reagent[A,B] {
  def tryReact(a: A, rx: Reaction, offer: Offer[B]) = 
    c(a).tryReact((), rx, offer)
}
\end{lstlisting}
\nocaptionrule
\caption{Excerpts from core reagent subclasses}
\end{figure}

@@

>>>>

%TODO: cite art of multiprocessor programming

@@

\begin{figure}
\begin{lstlisting}[frame=single]
class TreiberStack[A] {
  private val head = new AtomicReference[List[A]](Nil)
  def push(a: A) {
    val backoff = new Backoff
    while (true) {
      val cur = head.get()
      if (head.cas(cur, a :: cur)) return
      backoff.once()
    }
  }
  def tryPop(): Option[A] = {
    val backoff = new Backoff
    while (true) {
      val cur = head.get() 
      cur match {
        case Nil     => return None
        case a::tail => 
          if (head.cas(cur, tail)) return Some(a)
      }
      backoff.once()
    }
  }
}
\end{lstlisting}
\nocaptionrule
\caption{Treiber's stack (in Scala)}
\label{fig:classic-treiber}
\end{figure}

@@

\elide{
In \secref{background}, we saw a range of strategies for communication and
synchronization at a fine grain:
\begin{itemize}
\item using CAS to make direct, optimistic, and atomic updates;
\item using exponential backoff to ease contention; 
\item using side-channel communication to increase parallelism; 
\item using disjoint spinwaiting to eliminate useless cache traffic; 
\item using dual data structures to incorporate blocking.
\end{itemize}
Reagents incorporate these disparate techniques into a uniform set of
combinators for writing concurrent algorithms.  Some of the techniques are
built into the combinators themselves, while others are expressible as
reusable abstractions.
}

@@

// combinators for composition
+  : Reagent[A,B] * Reagent[A,B] => Reagent[A,B]
>> : Reagent[A,B] * Reagent[B,C] => Reagent[A,C]
|||  : Reagent[A,B] * Reagent[A,C] => Reagent[A,(B,C)]

// combinators for computation
ret: A => Reagent[Unit, A]
computed: PartialFunction[A, Reagent[Unit, B]]
          => Reagent[A,B]
postCommit: (A => Unit) => Reagent[A,A]
\end{lstlisting}
\nocaptionrule
\caption{The Reagent API}
\label{fig:combinators}
\end{figure}

@@

Memory is shared between reagents using the type \lstinline{Ref[A]} of
atomically-updatable references.\footnote{ These references are only
  accessible through reagent combinators, which allows them to provide
  functionality beyond the built-in \lstinline{AtomicReference[A]} type, such
  as the blocking logic we will see shortly.  }
The \lstinline{upd} combinator~(\figref{combinators}) represents simple atomic
updates to references.  It takes an \emph{update function}, which tells how to
transform a snapshot of the reference cell and some input into an updated
value for the cell and some output.  Using \lstinline{upd}, we can rewrite
\lstinline{TreiberStack} in a more readable and concise way:
\begin{lstlisting}
class TreiberStack[A] {
  private val head = new Ref[List[A]](Nil)
  val push: Reagent[A, Unit] = upd(head) {
    case (xs, x) => (x:xs, ())
  }
  val tryPop: Reagent[Unit, Option[A]] = upd(head) {
    case (x::xs, ()) => (xs, Some(x))
    case (Nil,   ()) => (Nil, None)
  }
}
\end{lstlisting}
In Scala, anonymous partial functions are written as a series of cases
enclosed in braces.  For \lstinline{push} and \lstinline{tryPop}, the case
analyses are exhaustive, so the update functions are in fact total.
\lstinline{Unit} is akin to \lstinline{void}: it is a type with a single
member, written \lstinline{()}.

@@

The \lstinline{!} operator invokes a reagent as a reactant: the invocation
lasts for a single reaction, and any messages sent by the reagent are
consumed.  But sometimes it is useful for reagents to persist beyond a single
reaction, \emph{i.e.}, to act as catalysts.  For example, the following
function creates a catalyst that merges input from two endpoints and sends the
resulting pairs to another endpoint:
\begin{lstlisting}
def zip(in1: Endpoint[Unit,  A], 
        in2: Endpoint[Unit,  B], 
        out: Endpoint[(A,B), Unit]) = 
  dissolve((swap(in1) ||| swap(in2)) >> swap(out))
\end{lstlisting}
The \lstinline{dissolve} function takes a \lstinline{Reagent[Unit, Unit]} and
introduces it as a catalyst.\footnote{For simplicity, we have not given a way
  to cancel catalysts after they have been introduced, but cancellation is
  easy to add.}  Operationally, in this example, that just means sending
messages along \lstinline{in1} and \listinline{in2} that are marked as
catalyzing messages, and hence are not consumed during reaction.  The upshot
is that senders along \lstinline{in1} will see the catalyzing messages, look
for messages along \lstinline{in2} to pair with, and ultimately send messages
along \lstinline{out} (and similarly in the other order).

@@

\begin{figure}
\begin{lstlisting}[frame=single]
class MSQueue[A] {
  private case class Node(data: A, next: Ref[Node])
  private val head = new Ref(new Node(null))  // sentinel
  private val tail = new Ref(read(head) ! ()) // sentinel
  val tryDeq: Reagent[Unit, Option[A]] = upd(head) {
    case (Node(_, Ref(n@Node(x, _))), ()) => (n, Some(x))
    case (emp,                        ()) => (emp, None)
  }
  private def findAndEnq(n: Node): Reagent[Unit,Unit] = 
    read(tail) ! () match {
      case ov@Node(_, r@Ref(null)) => 
        cas(r, null, n) >>
        postCommit { _ => cas(tail, ov, n) !? () }
      case ov@Node(_, Ref(nv)) => 
        cas(tail, ov, nv) !? (); 
        findAndEnq(n)
    }
  val enq: Reagent[A, Unit] = computed { 
    (x: A) => findAndEnq(new Node(x, new Ref(null)))
  }
}
\end{lstlisting}
\nocaptionrule
\caption{The Michael-Scott queue, with reagents}
\label{fig:msqueue}
\end{figure}

An example of this phenomenon is the classic Michael-Scott lock-free
queue~\cite{Michael1996}.  Unlike a stack, in which all activity focuses on
the head, queues have two loci of updates.  That means, in particular, that
the \lstinline{Ref}s used by its reagents may vary depending on the current
state of the queue.

Here is a brief overview of the Michael-Scott queue.  The queue is represented
as a \emph{mutable} linked list, with a sentinel node at the head (front) of
the queue.  The head pointer always points to the current sentinel node; nodes
are dequeues by a CAS to this pointer, just like Treiber stacks (but lagged by
one node).  The tail of the queue is the unique node, reachable from the head
pointer, with a null \lstinline{next} pointer; thanks to the sentinel, such a
node is guaranteed to exist.  If the queue is empty, the same node will be the
head (sentinel) and tail.  Finally, as an optimization for enqueing, a tail
\emph{pointer} is maintained with the invariant that the real tail is always
reachable from it.  The tail pointer may lag behind, however, which allows the
algorithm to work using only single-word CAS instructions.

Our reagent-based implementation of the Michael-Scott queue is shown
in~\figref{msqueue}.  The node representation is given as an inner \emph{case}
class.  In Scala, case classes provide two features we take advantage of.
First, the parameters to their constructors (here \lstinline{data} and
\lstinline{next}) are automatically added as final fields to the class, which
are initialized to the constructor argument values.  Second, they extend
pattern matching through \lstinline{case} so that instances can be
\emph{deconstructed}.  A pattern like \lstinline{case Node(d, n)} matches any
instance of the node class, binding \lstinline{d} to its \lstinline{data}
field and \lstinline{n} to its \lstinline{next} field.

The \lstinline{head} and \lstinline{tail} references of the queue are
initialized to the same sentinel node.  Here we use a new reagent,
\lstinline{read}~(\figref{combinators}), to extract the sentinel value from
the head when constructing the tail.  The read reagent in \lstinline{tail}'s
initializing expression is immediately executed during construction of an
\lstinline{MSQueue} instance.

The \lstinline{tryDeq} reagent is very similar to the \lstinline{tryPop}
reagent in \lstinline{TreiberStack}, modulo the sentinel node.  The reagent
pattern matches on the sentinel node, ignoring its \lstinline{data} field by
using \lstinline{_}, the wildcard.  The \lstinline{next} field is then matched
to a nested pattern, \lstinline{Ref(n@Node(x, _))}.  This pattern immediately
reads the current value of the reference stored in \lstinline{next}, binds
that value to \lstinline{n}, and then matches the
pattern~\lstinline{Node(x,_)} against \lstinline{n}.  If the pattern
matches---which it will any time the \lstinline{next} field of the sentinel is
non-null---the node \lstinline{n} becomes the new \lstinline{head} (and hence
the new sentinel).  
%TODO: picture?

Since the location of the tail node is determined dynamically by the data in
the queue, the \lstinline{enq} reagent must itself be determined dynamically.

The \lstinline{enq} reagent, on the other hand, must work with the shifting
position of the tail, which requires new combinators.

The \lstinline{cas} combinator~(\figref{combinators}) takes a
\lstinline{Ref[A]} and two \lstinline{A} arguments, giving the expected and
updated values, respectively.  Unlike its counterpart for
\lstinline{AtomicReference}, a \lstinline{cas} reagent does \emph{not} yield a
boolean result.  Instead, the reagent library builds in handling of failed
CASes, including an optimistic retry loop and exponential backoff.  A failure
to CAS is, as discussed in \secref{choice}, a transient failure that causes
the reagent to retry.

%TODO: treat !? as shorthand?

@@

\elide{ The \lstinline{read} combinator has a straightforward signature: if
  \lstinline{r} has type \lstinline{Ref[A]}, then \lstinline{read(r)} is a
  \emph{value} of type \lstinline{Reagent[Unit, A]} which, when invoked, takes
  a unit argument and yields an \lstinline{A} result.
}

@@

@@

\subsection{Caveats for composition}

@@

\begin{figure}
\begin{lstlisting}[frame=single]
def !(a: A): B = {
  val backoff = new Backoff
  def withoutOffer(): B =
    tryReact(a, empty, null) match {
      case Block => withOffer()
      case Retry => 
        backoff.once()
        if (maySync) withOffer() else withoutOffer()
      case ans   => ans.asInstanceOf[B]
    }
  def withOffer(): B = {
    val offer = new Offer[B]
    tryReact(a, empty, offer) match {
      case (f: Failure) => 
        if (f == Block) park() else backoff.once()
        if (offer.rescind) withOffer() else offer.answer
      case ans => ans.asInstanceOf[B]
    }
  }
  withoutOffer()
}
\end{lstlisting}
\nocaptionrule
\caption{The \lstinline{!} method, defined in \lstinline{Reagent[A,B]}}
\label{fig:reaction}
\end{figure}

@@

In the implementation, \lstinline{Reagent[A,B]} is an abstract class all of
whose subclasses are hidden.  The subclasses roughly correspond to the
combinator functions (which are responsible for instantiating them), and
instances of the subclasses store the arguments given to their corresponding
combinator.  Each subclass provides an implementation of the
\lstinline{tryReact} method, which is an abstract method of
\lstinline{Reagent[A,B]} with the following signature:\footnote{ Functional
  programming aficionados should regard \lstinline{tryReact} as a kind of
  Kleisli arrow for reagents.}
\begin{lstlisting}
  def tryReact(a: A, rx: Reaction, offer: Offer[B]): Any
\end{lstlisting}
The \lstinline{Any} type in Scala lies at the top of the subtyping hierarchy,
akin to \lstinline{Object} in Java.  Here we are using \lstinline{Any} to
represent a union of the type \lstinline{B} with the type
\lstinline{Failure}, where the latter has just two singleton instances,
\lstinline{Block} and \lstinline{Retry}, corresponding to permanent and
transient failures.

@@

\subsection{The entry point: reacting}
\label{sec:impl-reaction}
%TODO: note tail recursion

@@

\begin{figure}
\begin{lstlisting}[frame=single]
class Commit[A] extends Reagent[A,A] {
  def tryReact(a: A, rx: Reaction, offer: Offer[A]) = 
    if (offer != null && !offer.rescind) offer.answer
    else if (rx.commit) a 
    else Retry
}
\end{lstlisting}
\vskip -12pt
\begin{lstlisting}[frame=single]
class CAS[A,B](ref: Ref[A], ov: A, nv: A, k: Reagent[A,B]) 
  extends Reagent[Unit,B] {
  def tryReact(u: Unit, rx: Reaction, offer: Offer[B]) = 
    if (!rx.hasCAS && !k.hasCAS)
      if (ref.cas(ov, nv))
        k.tryReact((), rx, offer)
      else Retry
    else
      k.tryReact((), rx.withCAS(ref, ov, nv), offer)
}
\end{lstlisting}
\vskip -12pt
\begin{lstlisting}[frame=single]
class Choice[A,B](r1: Reagent[A,B], r2: Reagent[A,B]) 
  extends Reagent[A,B] {
  def tryReact(a: A, rx: Reaction, offer: Offer[B]) = 
    r1.tryReact(a, rx, offer) match {
    case Retry => 
        r2.tryReact(a, rx, offer) match {
          case Retry => Retry
          case Block => Retry // must still retry r1
          case ans   => ans
        }
      case Block => r2.tryReact(a, rx, offer)
      case ans => ans
    }
}
\end{lstlisting}
\vskip -12pt
\begin{lstlisting}[frame=single]
class Computed[A,B](c: A => Reagent[Unit,B]) 
  extends Reagent[A,B] {
  def tryReact(a: A, rx: Reaction, offer: Offer[B]) = 
    c(a).tryReact((), rx, offer)
}
\end{lstlisting}
\nocaptionrule
\caption{Excerpts from core reagent subclasses}
\end{figure}

@@

\subsection{Catalysis}
